---
title: "California_Housing"
author: "Aditya Pansari"
date: "2024-02-02"
output:
  word_document: default
  html_document: default
---
### California Housing
### Let's see what are the pre-installed libraries in the system


```{r}
#  California Housing Prices 1990 

library(gdata)
library(plyr)
library(readxl)
library(dplyr)
library(tidyr)
library(gridExtra)
library(caTools) # for splitting the data
library(digest) # For hash functions
library(ggplot2) 
library(recipes)
library(caret)
library(data.table)
library(caret)
library(leaflet) # for map. visualization
library(sf)
library(viridis)
```
### Data Loading
```{r}
houses <- read.csv("housing.csv")
```
### Data Preprocessing / Taking  a Quick Look at the Data Structure
```{r}
summary(houses)
head(houses)
str(houses)
colnames(houses)
```
### Let's look if we have missing values
```{r}
missing_counts <- colSums(is.na(houses)) # counting missing values for each column
missing_counts
```
### Let's calculate the proportion of missing values in each column
```{r}
missing_proportion <- houses %>%
  summarise_all(~ mean(is.na(.))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "ProportionMissing")
missing_proportion
```

```{r}
median_total_bedrooms <- median(houses$total_bedrooms, na.rm = TRUE)
houses$total_bedrooms[is.na(houses$total_bedrooms)] <- median_total_bedrooms
```
### A vizualization of the distribution of bedrooms
```{r}

total_bedrooms <- ggplot(houses, aes(x = total_bedrooms)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(title = "",
       x = "Total Bedrooms",
       y = "") +
  theme_minimal()
total_bedrooms


total_bedrooms2 <- ggplot(houses, aes(x = total_bedrooms)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(title = "",
       x = "Total Bedrooms",
       y = "") +
  theme_minimal()
total_bedrooms2
```
### let's vizualize each numeric column with histograms to see the distribution of its data
```{r}
par(mfrow=c(3, 3)) # allows us to insert several graphs in one

hist(houses$longitude, main="Longitude", xlab="Value")
hist(houses$latitude, main="Latitude", xlab="Value")
hist(houses$housing_median_age, main="Housing Median Age", xlab="Value")
hist(houses$total_rooms, main="Total Rooms", xlab="Value")
hist(houses$total_bedrooms, main="Total Bedrooms", xlab="Value")
hist(houses$population, main="Population", xlab="Value")
hist(houses$households, main="Households", xlab="Value")
hist(houses$median_income, main="Median Income", xlab="Value")
hist(houses$median_house_value, main="Median House Value", xlab="Value")
```

```{r}
numeric_columns <- c(
  "longitude",
  "latitude",
  "housing_median_age",
  "total_rooms",
  "total_bedrooms",
  "population",
  "households",
  "median_income",
  "median_house_value"
)
hists <- list()

for (column in numeric_columns) {
  p <- ggplot(houses, aes(x = !!sym(column))) +
    geom_histogram(fill = "skyblue", col="blue") +
    labs(title = paste("Distribution of", column),
         x = column,
         y = "") +
    theme_minimal() 
  hists[[column]] <- p
}

histograms <- grid.arrange(grobs = hists, ncol = 3)

```
### Creating a test set
```{r}
# Let's split the dataset into training and testing sets (80% train, 20% test)
set.seed(123)  #to make this output identical at every run

#Option 1 
training_data <- sample.split(houses$median_house_value, SplitRatio = 0.8)

# Creating the training and test sets
train_set <- houses[training_data, ]
test_set <- houses[!training_data, ]

nrow(train_set)  #number of rows for each set
nrow(test_set)   

summary(train_set)
summary(test_set)
```

```{r}
# Creating test set. Option 2
# We need to define a function to check whether an instance should be in the test set, but before we need to find an identifier. 
# Since we do not have index or ID column, I will create ID column by combining latitude and longtitude columns together to make them identifiers for test_set_check function.  


houses$ID <- paste(houses$latitude, houses$longitude, sep = "_")
```

```{r}
set.seed(42)

# Let's use createDataPartition function to split the data based on the identifier column, 
# it tries to ensure that each partition (train or test) has a representative sample of the unique values in the identifier column (ID)

index <- createDataPartition(houses$ID, p = 0.7, list = FALSE) # Creating an index for splitting the data (70% training, 30% test)

# Training and splitting the data by ID column
train_set2 <- houses[index, ]
test_set2 <- houses[-index, ]

summary(train_set2)
head(train_set2)
```
# Let's create a histogram for visualizig median income because some experts assume that this attribute is important in predicting median housing prices
```{r}
med_income_hist <- ggplot(houses, aes(x = median_income)) +
  geom_histogram(fill = "skyblue", color = "black", bins = 30) +
  labs(title = "",
       x = "",
       y = "") +
  theme_minimal()
med_income_hist
```
```{r}
# Most median income values are clustered around 1 to 6 (Keep in mind that this values are capped. So 1-6 are basically $10,000–$60,000),  
# but some median incomes go far beyond 6. It is important to have a sufficient number of instances in our dataset for each stratum, 
# This means that we should not have too many strata, and each stratum should be large enough. 
# We will create an income category attribute with 5 categories (labeled from 1 to 5): category 1 ranges from 0 to 1.5 
# (i.e., less than $15,000), category 2 from 1.5 to 3, and so on:

houses$income_cat <- cut(houses$median_income,
                         breaks = c(0, 1.5, 3.0, 4.5, 6.0, Inf),
                         labels = c(1, 2, 3, 4, 5))


# Get value counts for the 'income_cat' column
income_cat_counts <- table(houses$income_cat)
income_cat_counts
```
### let's make another hist with income categories
```{r}
med_income_cat_bar <- ggplot(houses, aes(x = factor(income_cat))) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(title = "Median income categories",
       x = "",
       y = "") +
  theme_minimal()
med_income_cat_bar
```

```{r}
# So now we can see that the most median income is in the category 3 which ranges from 3 to 4.5 (or in other words median income between $30k to $45k. Do not forget that it is in 1990!)
# Now we are ready to do stratified sampling based on the income category.

set.seed(42)  # Set a random seed for reproducibility
split <- createDataPartition(houses$income_cat, p = 0.2, list = FALSE)

# We will create another stratified training and test sets
strat_train_set <- houses[split, ]
strat_test_set <- houses[-split, ]

# let's check the distribution of income categories in the test set
income_cat_prop <- prop.table(table(strat_test_set$income_cat))

# I want to arrange the proportions in descending order
income_cat_props_desc <- income_cat_prop[order(-income_cat_prop)]
income_cat_props_desc
```
### Let's find proportion of income categories on the whole dataset to compare how stratified sampling is different from random sampling.
```{r}
income_cat_prop_full <- prop.table(table(houses$income_cat)) # the proportions of income categories in the whole dataset
income_cat_prop_full

# To compare between stratified and random sampling, we need to ensure that income_cat column exists in test_set2 too
test_set$income_cat <- cut(test_set$median_income,
                            breaks = c(0, 1.5, 3.0, 4.5, 6.0, Inf),
                            labels = c(1, 2, 3, 4, 5))


random_income_cat_prop <- prop.table(table(test_set$income_cat)) # the proportions of income categories in the random test set
random_income_cat_prop

stratified_income_cat_prop <- income_cat_props_desc # the proportions of income categories in the stratified test set
stratified_income_cat_prop
```
### Let's create a data frame to compare the proportions
```{r}
compare_prop <- data.frame(
  "Overall" = income_cat_prop_full,
  "Random" = random_income_cat_prop,
  "Stratified" = stratified_income_cat_prop
)
compare_prop
```
```{r}
# Now after exploring sampling bias comparison of stratified versus purely random sampling we should remove the income_cat attribute so the data is back to its original state:

houses <- subset(houses, select = -income_cat) # removing income_cat from houses

test_set2<- subset(test_set, select = -income_cat) # removing income_cat from test_set2

test_set <- subset(test_set, select = -income_cat)  # removing income_cat from test_set
```

```{r}
copy_strat_train_set <- strat_train_set
```
### We will visualize geographical data
```{r}

map <- ggplot(copy_strat_train_set, aes(x = longitude, y = latitude)) +
  geom_point(color = "skyblue", alpha = 0.4) +
  labs(title = "A geographical scatterplot of the data",
       x = "Longitude",
       y = "Latitude") +
  theme_minimal()
map
```
```{r}
library(ggplot2)
map2 <- ggplot(copy_strat_train_set, aes(x = longitude, y = latitude)) +
  geom_point(aes(size = population, color = median_house_value), alpha = 0.4) +
  scale_size_continuous(range = c(1, 15)) +
  scale_color_gradient(low = "yellow", high = "red") +
  labs(
    title = "California Housing Prices",
    x = "Longitude",
    y = "Latitude",
    size = "Population",
    color = "Median House Value"
  ) +
  theme_minimal()+
  guides(color = guide_colorbar(title = "Median House Value"))


map2
```
```{r}
numeric_attributes <- copy_strat_train_set[, sapply(copy_strat_train_set, is.numeric)]
```
```{r}
correlation_matrix <- cor(numeric_attributes)
```

```{r}
median_house_value_corr <- correlation_matrix[,"median_house_value"]
sorted_correlation <- sort(median_house_value_corr, decreasing = TRUE)
```

```{r}
corr_df <- data.frame(Attribute = names(sorted_correlation), Correlation = sorted_correlation)
corr_plot <- ggplot(corr_df, aes(x = reorder(Attribute, Correlation), y = Correlation)) +
  geom_bar(stat = "identity", fill = "skyblue", alpha = 1.5) +
  coord_flip() +
  labs(
    title = "Correlation with Median House Value",
    x = "Attribute",
    y = "Correlation"
  ) +
  theme_minimal()

corr_plot
```

```{r}
corr_plot2 <- ggplot(copy_strat_train_set, aes(x = median_income, y = median_house_value)) +
  geom_point(color = "#fdbb84", alpha = 0.7) +
  labs(
    title = "Median Income vs. Median House Value",
    x = "Median Income",
    y = "Median House Value"
  ) +
  theme_minimal()

corr_plot2
```

```{r}
ggplot(houses,aes(x=houses$housing_median_age, y=houses$median_income))+
  geom_point()
```
```{r}
ggplot(houses,aes(x=houses$housing_median_age, y=houses$median_income))+
  geom_violin()
```

